Deriving Hamiltonian mechanics from physics, without passing through Lagrangian.

### 2014-06-23 2014ClassicalHamiltonianDerivation/classicalHamiltonianDerivation.tex
Highlight terms in their definitions. Few other minor improvements.

\title{Deriving Hamiltonian mechanics from physics, without passing through Lagrangians.}
\author{Gabriele Carcassi}
\affiliation{Brookhaven National Laboratory, Upton, NY 11973}
\email{carcassi@bnl.gov}
\date{August 25, 2013}

\begin{abstract}
We derive the Hamiltonian formulation of classical mechanics directly, without reference to Lagrangian mechanics.
We start from the definition of states and the labels used to identify them, and show how simple properties required
on those labels lead to the Hamiltonian formulation.\end{abstract}

@@

\begin{defn}\label{combine label}
Let $I_1$ and $I_2$ be two sets of labels. We call $I = \langle I_1, I_2 \rangle$ a combined set of labels if an injective map exists such that $\forall i \in I \rightarrowtail i_1 \in I_1, i_2 \in I_2$.
\end{defn}

\begin{defn}\label{orth}
The two label sets $I_1$ and $I_2$ are said to be orthogonal iff $I = \langle I_1, I_2 \rangle = I_1 \times I_2$.
\end{defn}

@@

\begin{defn}\label{statedef}
Let $\mathbbm{S}_I$ be the set of all possible configuration states of the infinitesimal subdivision. We call this set phase space. We call each element $\mathbbm{s}_i$ a state.
\end{defn}

@@

\textbf{State mapping}. We will now concentrate ourselves on deterministic and reversible evolution.
\begin{assump}\label{detAss}
(\emph{Determinism and Reversibility}) The system undergoes deterministic (future state identified by the present state) and reversible (past state identified by the present state) evolution.
\end{assump}

@@

\begin{defn}\label{labelsIndep}
Let $I_1$ and $I_2$ be two orthogonal sets of labels and $I = I_1 \times I_2$. Let $f:I \leftrightarrow I$ be a bijective map. The labels are said independent iff we can define two sets of orthogonal labels $I_1'$ and $I_2'$ such that we can define separate maps:
\begin{enumerate}
\item $I_1' \times I_2' = I$
\item there exists a bijective map $f_1: I_1 \leftrightarrow I_1'$
\item there exists a bijective map $f_2: I_2 \leftrightarrow I_2'$
\item $f(i) = \langle f_1(i_1), f_2(i_2)\rangle$
\end{enumerate}
\end{defn}

@@

\begin{defn}\label{disclabelsoverr}
Consider a continuous numeric range. We divide the full range into contiguous cells. Let $I$ be the set of cells. For each cell we have a center value $x: I \mapsto \mathbb{R}$ and a width $w: I \mapsto \mathbb{R}$. $I$ is a set of discrete labels over a continuous range.
\end{defn}

@@

\begin{defn}\label{discreteHomogeneous}
A set $I$ of discrete labels over a continuous range is said to be homogeneous if $w(i)=k$: the bins are of equal width.
\end{defn}

@@

\begin{defn}\label{continuousLabels}
A state variable $X$ is the continuous limit of a set $I$ of discrete labels over a continuous range.
\end{defn}

@@

\begin{defn}\label{sdof}
A degree of freedom is a label set $X$ given by a pair of homogeneous and orthogonal state variables $P$ and $Q$. We call these conjugate variables. A state variable is said conjugate of the first iff together they form a pair of conjugate variables.
\end{defn}

@@

This is the equivalent of \ref{widthMapping} for two state variable. The density $\rho(p,q)$ will be defined on cells of infinitesimal area proportional to $dq \wedge dp$. When mapping one cell to another, the infinitesimal area will remain the same. The area in phase space of one degree of freedom can then be used as a measure for the cardinality of the labels. \ref{labelsCount} becomes area conservation for conjugate variables.

\begin{prop}\label{sdofInvariance}
Let $v$ and $w$ be two vectors defined on the tangent space of the manifold identified by two conjugate variables. Let
\begin{align*}
\omega_{\alpha, \beta} = \left[
  \begin{array}{cc}
    0 & 1 \\
    -1 & 0 \\
  \end{array}
\right] \\
\end{align*}
then $v'^{\alpha} \omega_{\alpha, \beta} w'^{\beta}=v^{\alpha} \omega_{\alpha, \beta} w^{\beta}$ under a bijective map.
\end{prop}

@@

Simply expand \ref{genAntisim} with the metric defined in \ref{sdofInvariance}. We recognise Hamilton's equations for one degree of freedom.

@@

\begin{defn}\label{mdof}
Two degrees of freedom are said independent if the corresponding label set $X^1$ and $X^2$ are independent.
\end{defn}

\begin{prop}\label{mdofInvariance}
Let $v$ and $w$ be two vectors defined on the tangent space of the manifold identified by two independent degrees of freedom. Let $\alpha$ and $\beta$ be indexes for the state variables $q^1, p^1, q^2, p^2$. Let
\begin{align*}
\omega_{\alpha, \beta} =  \left[
  \begin{array}{cc}
    1 & 0 \\
    0 & 1 \\
  \end{array}
\right] \otimes \left[
  \begin{array}{cc}
    0 & 1 \\
    -1 & 0 \\
  \end{array}
\right] =
\left[
  \begin{array}{cccc}
    0 & 1 & 0 & 0 \\
    -1 & 0 & 0 & 0 \\
    0 & 0 & 0 & 1 \\
    0 & 0 & -1 & 0 \\
  \end{array}
\right] \\
\end{align*}
then $v'^{\alpha} \omega_{\alpha, \beta} w'^{\beta}=v^{\alpha} \omega_{\alpha, \beta} w^{\beta}$ under a bijective map.
\end{prop}

The orthogonality of the labels corresponds to orthogonality in phase space, which the mapping needs to preserve across independent degrees of freedom. The cardinality of labels within each degree of freedom corresponds to the area in phase space\footnote{We assume we are using the same unit across d.o.f.}, which the mapping also needs to preserve. This is equivalent to require the conservation of the scalar product across independent degrees of freedoms, while still requiring conservation of the vector product within. That leads us to the metric defined by \ref{mdofInvariance}.
The metric generalizes \ref{sdofInvariance} to give us the cardinality of labels defined on the area given by two arbitrary vectors. For an infinitesimal region, this corresponds to $dq^1 \wedge dp^1 + dq^2 \wedge dp^2$, the sum of the projections on the independent planes.

@@

Expand \ref{genAntisim} with the metric defined in \ref{mdofInvariance}. We recognise Hamilton equations for multiple degrees of freedom.

@@

\begin{defn}\label{tdof}
The temporal degree of freedom is a label set $X$ given by the pair of conjugate variables $T$ and $E$. We call extended phase space the outer product between phase space and the temporal degree of freedom.
\end{defn}

@@

As in the case of independent d.o.f., we have three planes of conjugate variables forming a right triangle-like relationship with one side invariant.
\begin{align*}
&dq^1 \wedge dp^1 + dq^2 \wedge dp^2 = k \\
&dt \wedge de + k = dq \wedge dp
\end{align*}
In the previous case, the right angle was between the two independent degree of freedom. In this case, the right angle is between the invariant and the plane of constant $\langle Q, P \rangle$ where $dt \wedge de$ is defined. We rewrite it as $dq \wedge dp - dt \wedge de = k$. Which corresponds to the Minkowski product across d.o.f. and the vector product within. The metric, with a space-like convention, still gives us the cardinality of labels within a degree of freedom.\footnote{Adjusted to avoid double counting}

@@

Expand \ref{genAntisim} with the metric defined in \ref{tdofInvariant}, with the parameter $s$ instead of $t$ and generator $\mathcal{H}$ instead of $H$. We recognise Hamilton equations in the extended phase space.

It should not be a surprise that the equations do not mention $c$. In fact, nothing says that all $q^i$ represent space, and with the same unit, and that the laws of motion are invariant in all inertial frames. The only requirement we have is that the areas of each degrees of freedom represent the same cardinality for labels.

@@

Since $\mathcal{H}$ is constant through the evolution, it can serve both as the generating function and as the evolution constrain. By convention, we could set $\mathcal{H}=0$ without loss of generality as changing $\mathcal{H}$ by a constant does not change the equation of motion. This reduces extended phase space to $\mathbb{R}^{2*N + 1}$, the state labels plus time.

\textbf{Example}. We wrap up with an example. Let $\mathcal{H} = mc^2 + ((p^i)^2 - E^2/c^2) / 2m$.  If we apply \ref{tdofHam} we have:
\begin{align*}
d_{s}t &= E / mc^2 \\
d_{s}E &= 0 \\
d_{s}q^i &= p^i / m \\
d_{s}p^i &= 0
\end{align*}
For particles, let $s=\tau$:
\begin{align*}
E / c &= m c d_{\tau}t = m U^0 = P^0 \\
p^i &= m d_{\tau}q^i = m U^i = P^i \\
\end{align*}
so we recognize the four-momentum $P^\alpha = {E/c, p^i}$ and $\tau$ proper time. For anti-particles, let $s=-\tau$:
\begin{align*}
- E / c &= m c d_{\tau}t = m U^0 = P^0 \\
- p^i &= m d_{\tau}q^i = m U^i = P^i \\
\end{align*}
we end up with a minus sign between the four-momentum and the conjugate variables $P^\alpha = {-E/c, -p^i}$.\footnote{This is consistent with a free Dirac field ...} $\mathcal{H}$ relates to the rest energy, the parametrization to proper time (its conjugate). Parametrization and time are aligned for particles and anti-aligned for anti-particles.

@@

The hope is that, by continuing in this approach, we can shed more light on why nature has the laws it has; and show that they are not some arbitrary rules, but indeed necessary given few simple assumptions.

>>>>

\title{Deriving Hamiltonian mechanics from physics, without passing through Lagrangian.}
\author{Gabriele Carcassi}
\affiliation{Brookhaven National Laboratory, Upton, NY 11973}
\email{carcassi@bnl.gov}
\date{May 23, 2014}

\begin{abstract}
We derive the Hamiltonian formulation of classical mechanics directly, without reference to Lagrangian mechanics.
We start from the definition of states in terms of labels used to identify them, and show how, under a deterministic and
reversible process, the conservation of the cardinality of the labels leads to Hamilton's equations.\end{abstract}

@@

\begin{defn}\label{combine label}
Let $I_1$ and $I_2$ be two sets of labels. We call $I = \langle I_1, I_2 \rangle$ a \emph{combined set of labels} if an injective map exists such that $\forall i \in I \rightarrowtail i_1 \in I_1, i_2 \in I_2$.
\end{defn}

\begin{defn}\label{orth}
The two label sets $I_1$ and $I_2$ are said to be \emph{orthogonal} iff $I = \langle I_1, I_2 \rangle = I_1 \times I_2$.
\end{defn}

@@

\begin{defn}\label{statedef}
Let $\mathbbm{S}_I$ be the set of all possible configuration states of the infinitesimal subdivision. We call this set \emph{phase space}. We call each element $\mathbbm{s}_i$ a \emph{state}.
\end{defn}

@@

\textbf{State mapping}. We will now concentrate on deterministic and reversible evolution.
\begin{assump}\label{detAss}
(\emph{Determinism and Reversibility}) The system undergoes deterministic (future state identified by the present state) and reversible (past state identified by the present state) evolution.
\end{assump}

@@

\begin{defn}\label{labelsIndep}
Let $I_1$ and $I_2$ be two orthogonal sets of labels and $I = I_1 \times I_2$. Let $f:I \leftrightarrow I$ be a bijective map. The labels are said \emph{independent} iff we can define two sets of orthogonal labels $I_1'$ and $I_2'$ such that we can define separate maps:
\begin{enumerate}
\item $I_1' \times I_2' = I$
\item there exists a bijective map $f_1: I_1 \leftrightarrow I_1'$
\item there exists a bijective map $f_2: I_2 \leftrightarrow I_2'$
\item $f(i) = \langle f_1(i_1), f_2(i_2)\rangle$
\end{enumerate}
\end{defn}

@@

\begin{defn}\label{disclabelsoverr}
Consider a continuous numeric range. We divide the full range into contiguous cells. Let $I$ be the set of cells. For each cell we have a center value $x: I \mapsto \mathbb{R}$ and a width $w: I \mapsto \mathbb{R}$. $I$ is a set of \emph{discrete labels over a continuous range}.
\end{defn}

@@

\begin{defn}\label{discreteHomogeneous}
A set $I$ of discrete labels over a continuous range is said to be \emph{homogeneous} if $w(i)=k$: the bins are of equal width.
\end{defn}

@@

\begin{defn}\label{continuousLabels}
A \emph{state variable} $X$ is the continuous limit of a set $I$ of discrete labels over a continuous range.
\end{defn}

@@

\begin{defn}\label{sdof}
A \emph{degree of freedom} is a label set $X$ given by a pair of homogeneous and orthogonal state variables $P$ and $Q$. We call these \emph{conjugate variables}. A state variable is said conjugate of the first iff together they form a pair of conjugate variables.
\end{defn}

@@

This is the equivalent of \ref{widthMapping} for two state variables. The density $\rho(p,q)$ will be defined on cells of infinitesimal area proportional to $dq \wedge dp$. When mapping one cell to another, the infinitesimal area will remain the same. The area in phase space of one degree of freedom can then be used as a measure for the cardinality of the labels. \ref{labelsCount} becomes area conservation for conjugate variables.

\begin{prop}\label{sdofInvariant}
Let $v$ and $w$ be two vectors defined on the tangent space of the manifold identified by two conjugate variables. Let
\begin{align*}
\omega_{\alpha, \beta} = \left[
  \begin{array}{cc}
    0 & 1 \\
    -1 & 0 \\
  \end{array}
\right] \\
\end{align*}
then $v'^{\alpha} \omega_{\alpha, \beta} w'^{\beta}=v^{\alpha} \omega_{\alpha, \beta} w^{\beta}$ under a bijective map.
\end{prop}

@@

Simply expand \ref{genAntisim} with the metric defined in \ref{sdofInvariant}. We recognise Hamilton's equations for one degree of freedom.

@@

\begin{defn}\label{mdof}
Two degrees of freedom are said \emph{independent} if the corresponding label set $X^1$ and $X^2$ are independent.
\end{defn}

\begin{prop}\label{mdofInvariant}
Let $v$ and $w$ be two vectors defined on the tangent space of the manifold identified by two independent degrees of freedom. Let $\alpha$ and $\beta$ be indexes for the state variables $q^1, p^1, q^2, p^2$. Let
\begin{align*}
\omega_{\alpha, \beta} =  \left[
  \begin{array}{cc}
    1 & 0 \\
    0 & 1 \\
  \end{array}
\right] \otimes \left[
  \begin{array}{cc}
    0 & 1 \\
    -1 & 0 \\
  \end{array}
\right] =
\left[
  \begin{array}{cccc}
    0 & 1 & 0 & 0 \\
    -1 & 0 & 0 & 0 \\
    0 & 0 & 0 & 1 \\
    0 & 0 & -1 & 0 \\
  \end{array}
\right] \\
\end{align*}
then $v'^{\alpha} \omega_{\alpha, \beta} w'^{\beta}=v^{\alpha} \omega_{\alpha, \beta} w^{\beta}$ under a bijective map.
\end{prop}

The orthogonality of the labels corresponds to orthogonality in phase space, which the mapping needs to preserve across independent degrees of freedom. The cardinality of labels within each degree of freedom corresponds to the area in phase space\footnote{We assume we are using the same unit across d.o.f.}, which the mapping also needs to preserve. This is equivalent to require the conservation of the scalar product across independent degrees of freedoms, while still requiring conservation of the vector product within. That leads us to the metric defined by \ref{mdofInvariant}.
The metric generalizes \ref{sdofInvariant} to give us the cardinality of labels defined on the area given by two arbitrary vectors. For an infinitesimal region, this corresponds to $dq^1 \wedge dp^1 + dq^2 \wedge dp^2$, the sum of the projections on the independent planes.

@@

Expand \ref{genAntisim} with the metric defined in \ref{mdofInvariant}. We recognise Hamilton equations for multiple degrees of freedom.

@@

\begin{defn}\label{tdof}
The \emph{temporal degree of freedom} is a label set $X$ given by the pair of conjugate variables $T$ and $E$. We call extended phase space the outer product between phase space and the temporal degree of freedom.
\end{defn}

@@

As in the case of independent d.o.f., we have right triangle-like relationship between the plane where the invariant is defined and its projections on the planes defined by each d.o.f.:
\begin{align*}
&dq^1 \wedge dp^1 + dq^2 \wedge dp^2 = k \\
&dt \wedge de + k = dq \wedge dp
\end{align*}
In the previous case, the right angle was between the two independent d.o.f.. In this case, the right angle is between the invariant and the plane of constant $\langle Q, P \rangle$ where $dt \wedge de$ is defined. We rewrite it as $dq \wedge dp - dt \wedge de = k$. Which corresponds to the Minkowski product across d.o.f. and the vector product within. The metric, with a space-like convention, still gives us the cardinality of labels within a degree of freedom.\footnote{Adjusted to avoid double counting}

@@

Take the metric from \ref{tdofInvariant}, add multiple independent d.o.f as in \ref{mdofInvariant}, use \ref{genAntisim} with the parameter $s$ instead of $t$ and generator $\mathcal{H}$ instead of $H$. We recognise Hamilton equations in the extended phase space.

It should not be a surprise that the equations do not mention the speed of light $c$. In fact, nothing says that all $q^i$ represent space, and with the same unit, and that the laws of motion are invariant in all inertial frames. The only requirement we have is that the areas of each degrees of freedom represent the same cardinality for labels.

@@

Since $\mathcal{H}$ is constant through the evolution, it can serve both as the generating function and as the evolution constrain. By convention, we could set $\mathcal{H}=0$ without loss of generality as changing $\mathcal{H}$ by a constant does not change the equation of motion. This reduces extended phase space to $\mathbb{R}^{2*N + 1}$, the state variables plus time.

\textbf{Example}. We wrap up with an example. Let $\mathcal{H} = mc^2 + ((p^i)^2 - E^2/c^2) / 2m$.  If we apply \ref{tdofHam} we have:
\begin{align*}
d_{s}t &= E / mc^2 \\
d_{s}E &= 0 \\
d_{s}q^i &= p^i / m \\
d_{s}p^i &= 0
\end{align*}
For particles, let $s=\tau$:
\begin{align*}
E / c &= m c d_{\tau}t = m U^0 = P^0 \\
p^i &= m d_{\tau}q^i = m U^i = P^i \\
\end{align*}
so we recognize the four-momentum $P^\alpha = [E/c, p^i]$ and $\tau$ proper time. For anti-particles, let $s=-\tau$:
\begin{align*}
- E / c &= m c d_{\tau}t = m U^0 = P^0 \\
- p^i &= m d_{\tau}q^i = m U^i = P^i \\
\end{align*}
we end up with a minus sign between the four-momentum and the conjugate variables $P^\alpha = [-E/c, -p^i]$.\footnote{This is consistent with the Dirac field where we have $[i\hbar\partial_t, -i\hbar\partial_{x^i}]$ for generators and $\psi^\dagger\gamma^0[i\hbar\partial_t, -i\hbar\partial_{x^i}]\psi$ for observables, and $\gamma^0$ is $1$ for particles and $-1$ for anti-particles.} $\mathcal{H}$ relates to the rest energy, the parametrization to proper time (its conjugate). Parametrization and time are aligned for particles and anti-aligned for anti-particles.

@@

The hope is that, by continuing in this approach, we can shed more light on why the laws of physics are what they are; and show that they are not arbitrary rules, but necessary given few simple assumptions.
